---
title: "Script1_Armbruster2026_QC-Filtering"
author: "Daniel Armbruster"
date: "2026-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


- In this script we loaded the mapped 10X Genomics data, converted it into a Seurat object, performed quality control, and filtered the dataset to retain cells with at least 500 expressed genes per cell, less than 8% mitochondrial content per cell, and positive GFP expression.

- The session info including the used packages and their versions is provided at the end of this script.


# 1. Load Packages

```{r, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(Seurat)
library(dplyr)
library(ggplot2)
```

# 2 Load Data and Generate Seurat Object

```{r}
# Load mapped output files into a sparse count matrix
allcells_sparse <- Read10X(data.dir = "Mapped_Data/")

# Generate Seurat object with an initial filtering for cells expressing > 200 genes
allcells <- CreateSeuratObject(allcells_sparse, min.features = 200)
```

# 3 Quality Control

## 3.1 Prepare Metadata for QC


```{r}
# Calculate log10-transformed genes per UMI ratio
allcells$log10GenesPerUMI <- log10(allcells$nFeature_RNA) / log10(allcells$nCount_RNA)
```

## 3.2 Add further info to metadata

-   Store EGFP and MT content per cells in metadata

```{r}
# Add EGFP gene expression percentage
allcells$EGFPratio <- PercentageFeatureSet(object = allcells, pattern = "EGFP")

# Add mitochondrial gene expression percentage
allcells$MTratio <- PercentageFeatureSet(object = allcells, pattern = "mt-*")

# Add cell IDs to metadata
allcells$cells <- rownames(allcells@meta.data)

#Create a sample colum
allcells$sample <- "GFP+_sort"

# Rename metadata columns 
allcells@meta.data <- allcells@meta.data %>% dplyr::rename(seq_folder = orig.ident, nUMI = nCount_RNA, nGene = nFeature_RNA)

```

```{r}
# Save unfiltered Seurat object 
save(allcells, file="RDS_files/Script1_Armbruster2026_unfiltered.rds")
```

```{r}
# Extract metadata for QC
metadata <- allcells@meta.data
```

## 3.3 Assessing QC metrics

### 3.3.1 Cell number

```{r}
# Plot of cell number
metadata %>%
ggplot(aes(x=sample, fill=sample)) +
geom_bar() +
theme_classic() +
theme(legend.position = "none") +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
theme(plot.title = element_text(hjust=0.5, face="bold")) +
ggtitle("Cell number") +
annotate("text", x = 1, y = length(rownames(metadata))/2, label = length(rownames(metadata)), size = 5, color = "black")
```

### 3.3.2 Counts per cell

```{r}
metadata %>%
ggplot(aes(x=sample, y=log10(nUMI), fill=sample)) +
geom_violin() +
theme_classic() +
theme(legend.position = "none") +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
theme(plot.title = element_text(hjust=0.5, face="bold")) +
ggtitle("UMIs per cell") +
geom_jitter(shape=16, position=position_jitter(0.2), cex=1) +
stat_summary(fun.y=mean, geom="point", size=30, color="red", shape=95)
```

### 3.3.3 Genes per cell

Vln Plot:

```{r}
metadata %>%
ggplot(aes(x=sample, y=log10(nGene), fill=sample)) +
geom_violin() +
theme_classic() +
theme(legend.position = "none") +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
theme(plot.title = element_text(hjust=0.5, face="bold")) +
ggtitle("Genes per cell") +
geom_jitter(shape=16, position=position_jitter(0.2), cex=1) +
stat_summary(fun.y=mean, geom="point", size=30, color="red", shape=95) 
```

### 3.3.4 Mitochondrial Content

```{r}
# Define borders
lines3 <- data.frame(Ref = c("MT mean"), vals = mean(metadata$MTratio))

#Plot mitochondiral content as density plot
metadata %>%
ggplot(aes(color=sample, x=MTratio, fill= sample)) +
geom_density(alpha = 0.2) +
theme_classic() +
geom_vline(data = lines3 , aes(xintercept = vals, colour = Ref)) + 
annotate("text", x = lines3$vals + 0.35, y = 5, label = round(lines3$vals,2), size = 5, color = "turquoise")
```

### 3.3.5 EGFPratio

```{r}
# Define borders
lines3 <- data.frame(Ref = c("EGFP mean"), vals = mean(metadata$EGFPratio))

#Plot EGFP content as density plot
metadata %>%
ggplot(aes(color=sample, x=EGFPratio, fill= sample)) +
geom_density(alpha = 0.2) +
theme_classic() +
geom_vline(data = lines3 , aes(xintercept = vals, colour = Ref)) +
annotate("text", x = lines3$vals + 0.35, y = 1, label = round(lines3$vals,2), size = 5, color = "darkgreen")
```

### 3.3.6 Detected UMIs vs detected genes

```{r}
# Visualize the detected UMIs versus detected genes colored by mitochondrial content
metadata %>%
ggplot(aes(x=nUMI, y=nGene, color=MTratio)) +
geom_point() +
scale_colour_gradient(low = "gray90", high = "green") +
stat_smooth(method=lm) +
scale_x_log10() +
scale_y_log10() +
theme_classic() +
geom_hline(yintercept = 500) +
facet_wrap(~sample)

```

# 4 Filtering

## 4.1 Cell level filtering

-   Filtering thresholds:

    -   nGene \>500

    -   Mtratio \< 8 (only cells with less then 8 % mitochondrial reads)
    
    -   Positive for EGFP expression 

```{r}
filtered_cells <- subset(x = allcells, subset = (nGene >=500 & MTratio < 8 & EGFPratio > 0))
```

## 4.2 Gene level filtering

-   Will remove genes that are expressed in less than 5 cells

```{r}
print(paste("Gene number before:", length(rownames(filtered_cells))))

# Only keep genes expressed in at least 5 cells
counts <- GetAssayData(object = filtered_cells, slot = "counts")
nonzero <- counts > 0
keep_genes <- Matrix::rowSums(nonzero) >=5
filtered_counts <- counts[keep_genes,]
filtered_cells <- CreateSeuratObject(filtered_counts, meta.data = filtered_cells@meta.data)

print(paste("Gene number after:", length(rownames(filtered_cells))))
```

# 5. Reassess QC

```{r}
metadata_clean <- filtered_cells@meta.data
```

### 5.1 Cell number after filtering

```{r}
# Plot of cell number after filtering
metadata_clean %>%
ggplot(aes(x=sample, fill=sample)) +
geom_bar(fill ="seagreen4") +
theme_classic() +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
theme(plot.title = element_text(hjust=0.5, face="bold")) +
ggtitle("NCells") +
annotate("text", x = 1, y = length(rownames(metadata_clean))/2, label = length(rownames(metadata_clean)), size = 5, color = "black")
```

### 5.2 Counts per cell after filtering

```{r}
metadata_clean %>%
ggplot(aes(x=sample, y=log10(nUMI), fill=sample)) +
geom_violin(fill ="seagreen4") +
theme_classic() +
theme(legend.position = "none") +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
theme(plot.title = element_text(hjust=0.5, face="bold")) +
ggtitle("UMIs per cell") +
geom_jitter(shape=16, position=position_jitter(0.2), cex=1) +
stat_summary(fun.y=mean, geom="point", size=30, color="red", shape=95)
```

### 5.3 Genes per cell after filtering

```{r}
metadata_clean %>%
ggplot(aes(x=sample, y=log10(nGene), fill=sample)) +
geom_violin(fill ="seagreen4") +
theme_classic() +
theme(legend.position = "none") +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
theme(plot.title = element_text(hjust=0.5, face="bold")) +
ggtitle("Genes per cell") +
geom_jitter(shape=16, position=position_jitter(0.2), cex=1) +
stat_summary(fun.y=mean, geom="point", size=30, color="red", shape=95)
```

# 6 Save Filtered Seurat Object

```{r}
# Save filtered Seurat object
saveRDS(filtered_cells, file="RDS_files/Script1_Armbruster2026_filtered.rds")
```

```{r}
# Session info
sessionInfo()
```

```{r}
# Remove objects to reduce memory usage 
rm(list=ls())
gc()
```
