---
title: "Script2_Armbruster2026_doubletRemoval"
author: "Daniel Armbruster"
date: "2026-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- This script tries to identify and removes potential doublets from filtered RDS file we generated in Script 1.

- To this end we performed a preliminary Seurat processing workflow to make use of the DoubletFinder package. 

- The session info including the used packages and their versions is provided at the end of this script.

# 1. Load Packages:

```{r, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(DoubletFinder)
library(dplyr)
```

# 2. Load Data

```{r}
filtered_cells <- readRDS(file = "RDS_files/Script1_Armbruster2026_filtered.rds") 
```

# 3. Pre-processing for Doublet Detection

```{r}
# To detect potential doublets first a standard pre-processing workflow will be performed
normalized_cells <- NormalizeData(filtered_cells)
normalized_cells <- FindVariableFeatures(normalized_cells)
normalized_cells <- ScaleData(normalized_cells)
normalized_cells <- RunPCA(normalized_cells)
normalized_cells <- FindNeighbors(normalized_cells, dims= 1:20)
normalized_cells <- FindClusters(normalized_cells)
normalized_cells <- RunUMAP(normalized_cells, dims = 1:20, seed.use = 42)
DimPlot(normalized_cells)

```

# 4. Doublet Finder

## 4.1 Calculation of pK

```{r}
sweep.res.list_nsclc <- paramSweep(normalized_cells, PCs = 1:20, sct = FALSE)
```

```{r}
sweep.stats <- summarizeSweep(sweep.res.list_nsclc, GT = FALSE)
```

- find.pK to evaluate BCmetric across different pK values to identify the optimal parameter for DoubletFinder.

```{r}
bcmvn <- find.pK(sweep.stats)
```

```{r}
ggplot(bcmvn, aes(pK, BCmetric, group = 1)) + 
  geom_point() +
  geom_line()
```

-   Seems that there is a maximum at 0.22

```{r}
pK <- 0.22
```

## 4.2 Expected Doublet Number

- Based on the targeted loading of approximately 10,000 cells on the 10X chip, we assume a conservative expected doublet rate of 8% (slightly above the estimated 7.6% rate reported by the 10X Genomics guidelines).

```{r}
annotations <- normalized_cells@meta.data$seurat_clusters
homotypic.prop <- modelHomotypic(annotations)
nExp_poi <- round(0.08*nrow(normalized_cells@meta.data)) 
nExp_poi.adj <- round(nExp_poi*(1-homotypic.prop))
```

## 4.3 Run doubletFinder

```{r}
# Run DoubletFinder using the optimized pK value and expected doublet number
normalized_cells <- doubletFinder(normalized_cells,
                                     PCs = 1:20,
                                     pN = 0.25,
                                     pK = pK,
                                     nExp = nExp_poi,
                                     sct = FALSE)
```

```{r}
# Visualize predicted singlets and doublets on the UMAP embedding:
DimPlot(normalized_cells, reduction = "umap", group.by = "DF.classifications_0.25_0.22_271")

# Count predicted singlets and doublets:
table(normalized_cells@meta.data$DF.classifications_0.25_0.22_271)
```


# 5 Filter for predicted singlets 

```{r}
# Retain only cells classified as singlets for downstream analysis
cells_singlets <- subset(x = normalized_cells, subset = DF.classifications_0.25_0.22_271 == "Singlet" )
```

# 6 Save singlet Seurat object

```{r}
# Save filtered singlet Seurat object
saveRDS(cells_singlets, file = "RDS_files/Script2_Armbruster2026_singlets.rds")
```

```{r}
# Session info
sessionInfo()
```


```{r}
# Remove objects to reduce memory usage 
rm(list=ls())
gc()
```
