---
title: "Script4_Armbruster2026_RajOwn_TFcodes"
author: "Daniel Armbruster"
date: "2026-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- This script produced the z-score heatmap of cluster-specific transcription factors for the CA clusters compared to the whole-brain reference as visualized in Figure 2A as well as the associated data tables presented in Table S2A-R 

- As the whole-brain reference, we used the 5 dpf zebrafish larval whole-brain dataset published by Raj and colleagues deposited as an RDS file on GEO under accession number GSE158142 (Raj et al., 2020; DOI: 10.1016/j.neuron.2020.09.023).

- To identify highly expressed transcription factors in the CA clusters relative to the whole-brain reference, we merged the datasets, normalized the combined data, and computed differentially expressed transcription factors (minimum two fold upregulation, maximum adjusted p-value of 0.05) for each CA cluster (presented in Figure 1C) compared with all whole-brain cells combined into a single reference cluster. Then we visualized the expression of the 50 most highly upregulated transcription factors for each cluster as z-scores.

- The session info including the used packages and their versions is provided at the end of this script.

# 1. Load Packages

```{r, message=FALSE, warning=FALSE}
library(Seurat)
library(dplyr)
library(openxlsx)
library(ggplot2)
library(readr)
library(AnnotationHub)
```

# 2. Load data and merge datasets
## 2.1 Prepare datasets for merging
```{r}
#### ----- Raj data ----- ####
# Read in the Raj et al. 5 dpf dataset
raj_5d_data <- readRDS("Resources/Raj2020_5dpf/GSE158142_zf5dpf_cc_filt.cluster.rds")

# Update the Seurat object to the current Seurat version
raj_5d_data <- UpdateSeuratObject(raj_5d_data)

# Read the feature annotation table containing Ensembl IDs and gene symbols
raj_features <- read.csv(
  "Resources/raj_features.tsv",
  sep = "\t",
  header = FALSE,
  col.names = c("ensembl.id", "gene.symbol")
)

# Rename metadata columns to match the format of the integrated dataset
raj_5d_data@meta.data <- raj_5d_data@meta.data %>% dplyr::rename(orig.cluster = res.1)


#### ----- Own data ----- ####
# Read in own data
own_data <- readRDS("RDS_files/Script3_Armbruster2026_annotated.rds")

# Read the feature annotation table containing Ensembl IDs and gene symbols
own_features = read.csv("Resources/own_features.tsv", sep="\t", header = F,col.names = c("ensembl.id", "gene.symbol", "column3"))  

# Subset  relevant clusters (only want to keep clusters representing brain populations)
own_data <- subset(own_data, idents = c("Hyp1","OB", "Hb1","Ama1", "Hb2","pTub", "Hyp2", "Hyp3", "PrT", "PN", "Hb3", "Ama2", "NEC", "Cb", "DMX", "Ama3" ))

# Rename metadata columns to match the format of the integrated dataset
own_data@meta.data <- own_data@meta.data %>% dplyr::rename(orig.cluster = Fig1_annotation, cell_barcode = cells, pt.mito = MTratio)



#### ----- Extract count matrices and metadata, then convert gene symbols to Ensembl IDs ----- ####
# Both datasets were generated using the same zebrafish reference genome but different Ensembl releases. As gene symbols might differ, we will convert them back to Ensembl IDs prior merging. 

#### ----- Raj data ----- ####
# Extract the raw count matrix
raj_counts <- GetAssayData(raj_5d_data,assay = "RNA", slot = "counts")

# Extract meta data
raj_meta <- raj_5d_data@meta.data

# Replace gene symbols with Ensembl IDs
matched_ids_raj <- raj_features$ensembl.id[match(rownames(raj_counts), raj_features$gene.symbol)]
raj_counts <- raj_counts[!is.na(matched_ids_raj), ]
rownames(raj_counts) <- matched_ids_raj[!is.na(matched_ids_raj)]

#### Own: 
# Extract the raw count matrix
own_counts <- GetAssayData(own_data,assay = "RNA", slot = "counts")

# Extract metadata
own_meta <- own_data@meta.data

# Replace gene symbols with Ensembl IDs
matched_ids_own <- own_features$ensembl.id[match(rownames(own_counts), own_features$gene.symbol)]
own_counts <- own_counts[!is.na(matched_ids_own), ]
rownames(own_counts) <- matched_ids_own[!is.na(matched_ids_own)]

# Retain only genes shared between both datasets
common_genes <- intersect(rownames(raj_counts), rownames(own_counts))
raj_counts <- raj_counts[common_genes, , drop = FALSE]
own_counts <- own_counts[common_genes, , drop = FALSE]
own_counts <- own_counts[rownames(raj_counts), , drop = FALSE]

# Verify that both matrices have identical gene order
identical(rownames(raj_counts), rownames(own_counts))

# Create new Seurat objects from the harmonized count matrices and metadata
own_seurat <- CreateSeuratObject(counts = own_counts, meta.data = own_meta, project = "own_data")
raj_seurat <- CreateSeuratObject(counts = raj_counts, meta.data = raj_meta, project = "raj_data")
```

## 2.2 Merge datasets and restore gene symbols

```{r}
# Merge datasets
merged_seurat <- merge(
  raj_seurat,
  y = own_seurat,
  add.cell.ids = c("raj", "own"),
  project = "5dpf_merged"
)

#Create a mapping table to convert Ensembl IDs back to gene symbols
ensembl_to_own <- setNames(own_features$gene.symbol,
                           own_features$ensembl.id)

# Map Ensembl IDs from the merged Seurat object back to gene symbols
mapped_symbols <- unname(ensembl_to_own[rownames(merged_seurat)])

# Check whether unmapped Ensembl IDs were generated
sum(is.na(mapped_symbols))

# Check whether mapping resulted in duplicated gene symbols
dup_symbols <- mapped_symbols[duplicated(mapped_symbols)]
length(unique(dup_symbols))

# No unmapped IDs or duplicated gene symbols were generated

# Replace Ensembl IDs with gene symbols as row names
rownames(merged_seurat) <- mapped_symbols
```

```{r}
# Remove objects to reduce memory usage 
rm(zf5d_old, own_data, own_counts, own_features, own_meta, raj_counts, raj_features, raj_meta, raj_seurat, own_seurat)
gc()
```


# 3. Identify differentially expressed transcription factors
## 3.1 Normalize and prepare cluster identities

```{r}
# Set original cluster annotations as identities for downstream differential expression analysis
Idents(object = merged_seurat) <- "orig.cluster"

# Normalize the merged dataset using SCTransform
merged_seurat <- SCTransform(merged_seurat)

# Prepare the SCT assay for differential expression testing
merged_seurat <- PrepSCTFindMarkers(merged_seurat, assay = "SCT")

# Combine Raj whole-brain clusters into a single whole-brain reference cluster (WB)
# (other clusters correspond to the CA clusters and retain their original label)
merged_seurat <- RenameIdents(merged_seurat, "0" = "WB", "1" = "WB", "2" = "WB", "3" = "WB", "4" = "WB", "5" = "WB", "6" = "WB", "7" = "WB", "8" = "WB", "9" = "WB", "10" = "WB", "11" = "WB", "12" = "WB", "13" = "WB", "14" = "WB", "15" = "WB", "16" = "WB", "17" = "WB", "18" = "WB", "19" = "WB", "20" = "WB", "21" = "WB", "22" = "WB", "23" = "WB", "24" = "WB", "25" = "WB", "26" = "WB", "27" = "WB", "28" = "WB", "29" = "WB", "30" = "WB", "31" = "WB", "32" = "WB", "33" = "WB", "34" = "WB", "35" = "WB", "36" = "WB", "37" = "WB", "38" = "WB", "39" = "WB", "40" = "WB", "41" = "WB", "42" = "WB", "43" = "WB", "44" = "WB", "45" = "WB", "46" = "WB", "47" = "WB", "48" = "WB", "49" = "WB", "50" = "WB", "51" = "WB", "52" = "WB")

merged_seurat$DEG_sample_names <- Idents(merged_seurat)

```

## 3.1 Perform differential expression analysis of each CA cluster against the whole-brain reference
```{r}
# Define clusters to analyse in the DEG loop
clusters_for_DEG <- levels(merged_seurat$DEG_sample_names)

# Load zebrafish transcription factor annotation list used to extract transcription factors from DEG results ( List from Animal Transcription Factor Database 3.0 (AnimalTFDB3.9))
TFlist <- read_tsv(file = "Resources/DanioRerio_TF.tsv")

# Rename gene symbol column and remove unnecessary annotation columns
names(TFlist)[names(TFlist) == "Symbol"] <- "gene"
TFlist$Species <- NULL
TFlist$Ensembl <- NULL
TFlist$Family <- NULL
TFlist$Protein <- NULL
TFlist$`Entrez ID` <- NULL

# Remove duplicated transcription factor entries
TFlist <- TFlist[!duplicated(TFlist$gene),]

# Define output directory for DEG tables
savepath <- "DEG_tables/Script4_RajOwn_TFs/"


# Initialize lists for the storage of DEG results and transcription factor subsets
tf_top50_list <- list()
all_markers_list <- list()
tf_markers_list <- list()

# Perform the differential expression analysis on the SCT data slot of each CA cluster against whole-brain reference
for (cluster in clusters_for_DEG) {
  
  print(paste("Current cluster:", cluster))
  markers_wb_vs_CA <- FindMarkers(
    merged_seurat,
    ident.1 = cluster,
    ident.2 = "WB",
    min.pct = 0.2,
    logfc.threshold = 1,
    only.pos = T
  )
  

  # Add gene names and cluster identity to the DEG table
  markers_wb_vs_CA$gene <- row.names(markers_wb_vs_CA)
  try(markers_wb_vs_CA$cluster <- cluster)
  
 # Retain significantly enriched genes and rank them by average log2 fold change
  try(markers_wb_vs_CA <- markers_wb_vs_CA %>% 
    dplyr::filter(p_val_adj <= 0.05)
      %>% dplyr::arrange(desc(avg_log2FC)))

  # Extract transcription factors from the DEG list and store the top 50 TFs ranked by average log2 fold change
  markers_wb_vs_CA_TFs <- inner_join(markers_wb_vs_CA, TFlist, by = "gene")
  top_TF_genes_up <- markers_wb_vs_CA_TFs %>% dplyr::arrange(desc(avg_log2FC))
  tf_top50_list[[cluster]] <- top_TF_genes_up$gene[1:50]
  
  # Store all DEGs and transcription factor DEGs
  all_markers_list[[cluster]] <- markers_wb_vs_CA
  tf_markers_list[[cluster]] <- markers_wb_vs_CA_TFs

  
}

# Export top 50 transcription factors per cluster
write.csv(tf_top50_list, file= paste(savepath, "tf_genes_sorted_lists/Script4_RajOwn_DEG_tfgenelist_Top50.csv", sep = ""))

# Rename list entries to show the comparison performed
names(all_markers_list) <- paste0(names(all_markers_list), "_vs_WB_allgenes")
names(tf_markers_list) <- paste0(names(tf_markers_list), "_vs_WB_TFgenes")

# Combine and export the DEG tables
combined_list <- c(all_markers_list, tf_markers_list)
combined_list <- combined_list[order(names(combined_list))]
write.xlsx(combined_list, file = paste0(savepath, "Script4_RajOwn_cluster_vs_WB_DEGs.xlsx"))
```

-   Heatmap geneartion:

```{r}
# Load the saved top 50 transcription factors for all clusters and make them unique
tf_all_clusters_50 <- read.csv(paste(savepath, "tf_genes_sorted_lists/Script4_RajOwn_DEG_tfgenelist_Top50.csv", sep=""))[-1]
all_TFs <- unique(unlist(tf_all_clusters_50))

# Calculate average SCT-normalized expression of the transcription factors per cluster
avg <- AverageExpression(
  merged_seurat,
  features = all_TFs,
  group.by = "DEG_sample_names",     
  slot = "data"
)


# Scale TF expression values across clusters for visualization
avg_mat <- avg[[DefaultAssay(merged_seurat)]]
avg_scaled <- t(scale(t(avg_mat)))
avg_scaled[is.na(avg_scaled)] <- 0
lim <- max(abs(avg_scaled))
breaks <- seq(-lim, lim, length.out = 101)

# Generate the heatmap of transcription factor expression patterns across clusters as z-score
DEG_heatmap <- pheatmap::pheatmap(avg_scaled,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  clustering_method = "average",
  color = colorRampPalette(c("navy","white","firebrick3"))(100),
  breaks = breaks,
  silent = TRUE,
  width = 10,
  height = 10,
  fontsize_row = 3,
  
)

# Save heatmap as SVG file
print(DEG_heatmap)
svg(paste(savepath, "Script4_Figure2_RajOwn_DEG_TF_heatmap_Top50.svg", sep = ""), width = 10, height = 10)
grid::grid.draw(DEG_heatmap$gtable)
dev.off()

# Prepare associated data for export
avg_mat <- as.data.frame(avg_mat)
avg_mat <- avg_mat[, order(names(avg_mat), decreasing=F)]
avg_mat <- avg_mat %>% dplyr::select("WB", everything())
avg_scaled <- as.data.frame(avg_scaled)
avg_scaled <- avg_scaled[, order(names(avg_scaled), decreasing=F)]
avg_scaled <- avg_scaled %>% dplyr::select("WB", everything())
dataframe_names <- list("avg_SCT_expression" = avg_mat, "scaled_avg_SCT_values" = avg_scaled)

# Export average SCT-normalized expression values and z-score transformed values used to generate the heatmap
openxlsx::write.xlsx(dataframe_names, file = paste(savepath, "Script4_RajOwn_HeatmapTable.xlsx", sep = ""), rowNames=T)
```

```{r}
# Identify the 3 transcription factors with the highest z-score per clusters to label them in the heatmap.
label_genes <- lapply(colnames(avg_scaled), function(cl){

  genes <- rownames(avg_scaled)[
    apply(avg_scaled, 1, which.max) == which(colnames(avg_scaled)==cl)
  ]

  zvals <- avg_scaled[genes, cl]

  head(genes[order(zvals, decreasing=TRUE)], 3)

})
label_genes <- unique(unlist(label_genes))
label_genes
```
```{r}
# Session info
sessionInfo()
```


```{r}
# Remove objects to reduce memory usage 
rm(list=ls())
gc()
```
